---
title: Comparison of ranks compiled from S1 and ST sobol indices and coefficients
  (COEF) of corresponding elastinet models
output:
  pdf_document: default
  html_notebook: default
---

```{r echo=FALSE,include=FALSE}
library(tidyverse)
library(reshape2)
library(tidyselect)

blue <- "#27647B"
yellow <- "#E69F00"
red <- "#DA5526"
```


```{r echo=FALSE,include=FALSE}
min_coef <- read_csv("compiled_ranks/Lambda_min_COEF_RANKS.csv")
min_s1 <- read_csv("compiled_ranks/Lambda_min_S1_RANKS.csv")
min_st <- read_csv("compiled_ranks/Lambda_min_ST_RANKS.csv")
se_coef <- read_csv("compiled_ranks/Lambda_1se_COEF_RANKS.csv")
se_s1 <- read_csv("compiled_ranks/Lambda_1se_S1_RANKS.csv")
se_st <- read_csv("compiled_ranks/Lambda_1se_ST_RANKS.csv")

ranks <- min_coef %>% mutate(lambda = "min", method = "coef") %>% 
  bind_rows(  min_s1 %>% mutate(lambda = "min", method = "s1") ) %>% 
  bind_rows(  min_st %>% mutate(lambda = "min", method = "st") ) %>% 
  bind_rows(  se_coef %>% mutate(lambda = "1se", method = "coef") ) %>%
  bind_rows(  se_s1 %>% mutate(lambda = "1se", method = "s1") ) %>%
  bind_rows(  se_st %>% mutate(lambda = "1se", method = "st") )

ranks_m <- ranks %>% melt(id.vars = c("Variable","lambda","method")) %>% rename(alpha = variable, rank = value) %>% 
  mutate(alpha = ifelse(alpha == "a0",0.0, ifelse(alpha == "a025", 0.25, ifelse(alpha == "a05", 0.5, ifelse( alpha == "a075",0.75,1) ) )  ))
```


#### Examination of absolute ranks

Visualisation of ranks comparison between pairs of S1 and COEF, as well as ST and COEF.

Dataset overview
```{r compile_rank_dataset, echo=FALSE}
ranks_m
```



Examination of differences of ranks between S1 and COEF, by visualising their absolute ranks.
```{r fig.asp=0.6, fig.width=8, echo=FALSE}

ranks_m %>% filter(method %in% c("coef","s1")) %>% ggplot(aes(Variable,rank, group=method, fill=method)) +
  geom_bar(stat = "identity", position = position_dodge(), width = 0.65) +
  facet_grid(alpha~lambda) + 
  theme_bw() +
  theme(axis.text.x = element_text(vjust = 1, angle = 90)) + 
  scale_fill_manual(values=c(blue,yellow))

```

Examination of differences of ranks between ST and COEF, by visualising their absolute ranks.
```{r fig.asp=0.6, fig.width=8, echo=FALSE}

ranks_m %>% filter(method %in% c("coef","st")) %>% ggplot(aes(Variable,rank, group=method, fill=method)) +
  geom_bar(stat = "identity", position = position_dodge(), width = 0.65) +
  facet_grid(alpha~lambda) + 
  theme_bw() +
  theme(axis.text.x = element_text(vjust = 1, angle = 90)) + 
  scale_fill_manual(values=c(blue,yellow))

```


#### Examination of rank differences
Following part compares differences in ranks compiled by S1,ST and COEF methods. Coparison is done between following pairs: S1-COEF and ST-COEF.


Dataset overview
```{r compile_rank_diff_dataset, echo=FALSE, warning=FALSE}
first_sc_ds <- ranks_m %>% dcast(Variable + lambda ~ method + alpha, value.var = "rank") %>% 
  mutate(
    diff_s1c_0 = s1_0 - coef_0,
    diff_s1c_0.25 = s1_0.25 - coef_0.25,
    diff_s1c_0.5 = s1_0.5 - coef_0.5,
    diff_s1c_0.75 = s1_0.75 - coef_0.75,
    diff_s1c_1 = s1_1 - coef_1,
    diff_stc_0 = st_0 - coef_0,
    diff_stc_0.25 = st_0.25 - coef_0.25,
    diff_stc_0.5 = st_0.5 - coef_0.5,
    diff_stc_0.75 = st_0.75 - coef_0.75,
    diff_stc_1 = st_1 - coef_1
    ) %>% 
  select(Variable,lambda, starts_with("diff")) %>% 
  melt(id.vars = c("Variable","lambda")) %>% 
  separate(variable,sep = "_", into = c("diff","combination","alpha")) %>% 
  select(-diff)
  

first_sc_ds

```


Comparison of differences between S1 and COEF ranks
```{r fig.asp=0.6, fig.width=8, echo=FALSE}

first_sc_ds %>% filter(combination == "s1c") %>% ggplot(aes(Variable,value, fill=alpha)) +
  geom_bar(stat = "identity") +
  facet_grid(alpha~lambda) + 
  theme_bw() +
  theme(axis.text.x = element_text(vjust = 1, angle = 90))

```

Comparison of differences between ST and COEF ranks
```{r fig.asp=0.6, fig.width=8, echo=FALSE}

first_sc_ds %>% filter(combination == "stc") %>% ggplot(aes(Variable,value, fill=alpha)) +
  geom_bar(stat = "identity") +
  facet_grid(alpha~lambda) + 
  theme_bw() +
  theme(axis.text.x = element_text(vjust = 1, angle = 90))

```



#### Examination of rank variance along elasticnet method (alpha)

Dataset overview
```{r compile_rank_variance_dataset, echo=FALSE}
ranks_m
```

Visualization of variance along alpha path, with ranks estimation based on S1,ST and COEF.
```{r fig.asp=0.6, fig.width=8, echo=FALSE}
ranks_m %>% ggplot(aes(Variable, rank, color=method)) +
  geom_boxplot() +
  #geom_bar(stat = "identity", position = position_dodge(), width = 0.65) +
  facet_grid(method~lambda) + 
  theme_bw() +
  theme(axis.text.x = element_text(vjust = 1, angle = 90)) + 
  scale_color_manual(values=c(blue,yellow,red))

```




